---
title: "Getting started with autotool"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{autotool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
box::use(autotool[generate_tool, make_handler])
```

`autotool` generates an OpenAI / Anthropic / DeepSeek tool definition
from any R function by reading its signature and its documentation.
This vignette walks through the four common situations:

1. A base-R function whose docs ship in installed `.Rd` files.
2. A function from your own source code with roxygen comments.
3. Hiding implementation-detail arguments from the model.
4. Overriding what introspection cannot see (complex types).

It closes with a short adapter chapter showing how to slot the output
into [`ellmer`](https://ellmer.tidyverse.org/), the tidyverse LLM
client, which uses its own `type_*()` constructors instead of raw
JSON schemas.

## 1. Wrapping a base-R function

`stats::rnorm` has installed help; `autotool` pulls the title and
per-argument descriptions straight from there.

```{r}
tool_rnorm <- generate_tool(stats::rnorm)
tool_rnorm$`function`$name
tool_rnorm$`function`$description
tool_rnorm$`function`$parameters$properties$mean$description
unlist(tool_rnorm$`function`$parameters$required)
```

The `required` list is derived from "no default in `formals()`" — no
manual second list to keep in sync.

`mean` has default `0` (numeric), so it is typed as `"number"` and the
default propagates to the schema:

```{r}
tool_rnorm$`function`$parameters$properties$mean
```

`n` has no default and R has no static types, so it falls back to
`"string"`. Override that with `schemas` (shown in step 4) or supply a
better description with `descriptions`:

```{r}
tool_rnorm <- generate_tool(
  stats::rnorm,
  descriptions = list(n = "Number of random values to draw.")
)
tool_rnorm$`function`$parameters$properties$n
```

## 2. Wrapping a source-loaded function

When a function is loaded from a source file (via `source()`,
`sys.source()`, `devtools::load_all()`, or `box::use()`), its `srcref`
attribute records the file. `autotool` reads the `#'` comments above
the definition directly — no `roxygen2` dependency.

```{r}
tmp <- tempfile(fileext = ".R")
writeLines(
  c(
    "#' Multiply two numbers together.",
    "#'",
    "#' @param a First factor.",
    "#' @param b Second factor.",
    "multiply <- function(a, b = 1) a * b"
  ),
  tmp
)
env <- new.env()
sys.source(tmp, envir = env, keep.source = TRUE)

tool_multiply <- generate_tool(env$multiply, name = "multiply")
tool_multiply$`function`$description
tool_multiply$`function`$parameters$properties$a$description
```

## 3. Hiding credentials and other infra arguments

This is the headline feature. Suppose you want to expose a real trading
function to a model but keep its `api_key`, `base_url`, and retry
budget pinned on your side.

```{r}
fetch_quotes <- function(
  symbols,
  start,
  end = Sys.time(),
  api_key = NULL,
  base_url = "https://example.com",
  retries = 3L
) {
  return(list(symbols = symbols, start = start, end = end))
}

tool_quotes <- generate_tool(
  fetch_quotes,
  defaults = list(
    api_key  = "sk-not-shown-to-model",
    base_url = "https://example.com",
    retries  = 5L
  ),
  schemas = list(
    symbols = list(
      type = "array",
      items = list(type = "string"),
      description = "Ticker symbols, e.g. ['AAPL','NVDA']"
    ),
    start = list(
      type = "string",
      description = "ISO datetime floor, e.g. '2026-04-01T00:00:00Z'"
    ),
    end = list(
      type = "string",
      description = "ISO datetime ceiling. Default: now."
    )
  )
)
names(tool_quotes$`function`$parameters$properties)
```

The model sees only `symbols`, `start`, `end`. Build a handler once
with `make_handler()` and call it whenever the model dispatches:

```{r}
handle_fetch_quotes <- make_handler(
  fetch_quotes,
  defaults = list(
    api_key = "sk-not-shown-to-model",
    base_url = "https://example.com",
    retries = 5L
  )
)

# When the model emits a tool call, parse its arguments and call:
model_emitted <- list(symbols = c("AAPL", "NVDA"), start = "2026-04-01")
handle_fetch_quotes(model_emitted)
```

The pinned values are merged in before the function runs. The model
never sees the key. Plug the closure into your own handler registry
(e.g., `TOOL_HANDLERS <- list(fetch_quotes = handle_fetch_quotes)`)
and dispatch by name in your agent loop.

If you need post-processing — caching, result truncation, logging —
just write a regular 1-arg function instead of using `make_handler()`.
The factory is a convenience, not a requirement.

## 4. Overriding what introspection cannot see

R's lack of static types means `autotool` infers from default values
and falls back to `"string"` otherwise. For complex shapes —
arrays-of-objects, nested records, custom classes — supply a full
schema yourself via `schemas`:

```{r}
submit_trades <- function(decisions, memo) invisible(NULL)

tool_submit <- generate_tool(
  submit_trades,
  description = "Submit a batch of trade decisions for execution.",
  schemas = list(
    decisions = list(
      type = "array",
      items = list(
        type = "object",
        properties = list(
          ticker = list(
            type = "string",
            description = "Ticker symbol, e.g. 'AAPL'."
          ),
          action = list(
            type = "string",
            enum = list("buy", "sell", "hold"),
            description = "Trade direction."
          ),
          notional = list(
            type = "number",
            description = "Dollar amount; 0 for hold."
          )
        ),
        required = list("ticker", "action", "notional")
      )
    ),
    memo = list(
      type = "string",
      description = "Short summary of this cycle's actions."
    )
  )
)
str(tool_submit$`function`$parameters$properties$decisions, max.level = 2)
```

## 5. Plugging into ellmer

`autotool` returns the OpenAI / DeepSeek tool shape. That format goes
straight into any OpenAI-compatible endpoint (DeepSeek, Mistral,
local llama.cpp, vLLM, etc.).

[`ellmer`](https://ellmer.tidyverse.org/), the tidyverse LLM client,
is the exception — its `tool()` function uses its own `type_*()`
constructors rather than accepting a raw JSON schema. A 20-line
adapter bridges the two so you still get autotool's auto-extracted
descriptions while talking to ellmer.

```{r ellmer-adapter, eval = FALSE}
to_ellmer_type <- function(name, prop, required_set) {
  ctor <- switch(
    prop$type,
    string = ellmer::type_string,
    integer = ellmer::type_integer,
    number = ellmer::type_number,
    boolean = ellmer::type_boolean,
    stop("Unsupported type for ellmer adapter: ", prop$type)
  )
  return(ctor(prop$description, required = name %in% required_set))
}

as_ellmer_tool <- function(fn, tool_def) {
  props <- tool_def$`function`$parameters$properties
  required_set <- unlist(tool_def$`function`$parameters$required)
  args <- mapply(
    to_ellmer_type,
    names(props),
    props,
    MoreArgs = list(required_set = required_set),
    SIMPLIFY = FALSE,
    USE.NAMES = TRUE
  )
  return(ellmer::tool(fn, tool_def$`function`$description, arguments = args))
}

# Usage:
ellmer_tool <- as_ellmer_tool(stats::rnorm, generate_tool(stats::rnorm))
```

This is a deliberate one-direction adapter — it doesn't try to handle
the array-of-objects shapes from step 4, which ellmer's `type_*()`
helpers don't directly express. For those cases pick whichever client
you prefer: send the autotool schema raw to an OpenAI-compatible
endpoint, or hand-write the ellmer arguments.

## Type-inference reference

| Default value           | Inferred `type`                      |
|-------------------------|--------------------------------------|
| no default              | `"string"` (override via `schemas`)  |
| `"value"`               | `"string"` with `default`            |
| `c("a", "b", "c")`      | `"string"` with `enum`               |
| `TRUE` / `FALSE`        | `"boolean"`                          |
| `1L`                    | `"integer"`                          |
| `1.5` (or `1`, `0`)     | `"number"`                           |
| `list(...)`             | `"object"` (use `schemas` for shape) |
